CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-05-26 script/en/chapters/numbertheory.tex
*** empty log message ***

\newpage
%\hypertarget{Kap_3}
\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, besslinger@web.de, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\

@@

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established by Carl Friedrich Gauss
\index{Gauss Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorisation, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss Carl Friedrich}
(Apr 4, 1777$-$Feb 23, 1855):}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

%\begin{quote}
%{\em Carl Friedrich Gauss (30.4.1777$-$23.2.1855):}\\
%Mathematics is the queen of sciences and number theory is the queen of mathematics.
%\end{quote}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823$-$Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, 06.10.1831$-$12.02.1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter \ref{primesinmath}} for which numbers in the set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

{\bf Example:} \\
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that each operand is divisible by $7$: 
$69,993 = 70,000 – 7$. Therefore, the difference is also divisible by $7$.

@@

\newpage
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Seneca\footnotemark:}\\
    The way of theory is long, it is short and effective by examples.
}}
\end{center}
\addtocounter{footnote}{0}
\footnotetext{%
     Lucius Annaeus Seneca, philosophical writer and poet, 4 B.\:C. $-$ 65 A.\:D.}

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Kap_3_8}{}\label{patternsandstructures}

As mathematicians investigate the structure \index{Structure} $a *x \equiv b$ mod $m$ (see \hyperlink{kap_3_5_2}{chapter 3.5.2}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most common asymmetric 
cryptography procedure. Developed in 1978 by Ronald Rivest, Adi Shamir and Leonard Adleman, it can be used both for
signatures and for encryption. Cryptographers always associate this procedure with the abbreviation ``RSA''. However, 
please note that the abbreviation may be used in other fields to mean
something completely different.}$^,$\footnote{%
In literature and movies not only classical methods are used (like the
secret message solved by Sherlock Holmes in ``The Adventure of the Dancing
Men'' by Arthur Conan  Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner\index{Elsner 1999}, 1999
  (included in the CrypTool package\index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}}
procedure in the group $\mathbb{Z}_n^*$.

@@

{\bf 2nd comment:} \\
When it comes to signing messages, we perform the same operations but first use the secret key $d$, followed by the 
public key $e$. The RSA procedure can also be used to create digital signatures\index{Signatures!digital}, because: 
$$
M \equiv (M^d)^e{\rm ~(mod~}n).
$$

@@

The security of the RSA algorithm is based on the empirical observation
that factoring large numbers is a hard problem. A Module $n$ (typically,
1024 bit) can be easily constructed as the product of two large primes $p$,
$q$ (typically, 500$-$600 bit each), by calculating $n=pq$. However, it is
a hard problem to extract $p$, $q$ from $n$.  Without knowing $p$ or $q$,
the private key cannot be calculated.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{The TWIRL Device} \label{TWIRLDevice} \index{TWIRL Device}\mbox{}

@@

The  computations for the factorization of RSA160 took place at the 
German Information Security Agency (GISA) in Bonn\footnote{%
Every year the GISA\index{GISA} creates a paper to describe which 
crypto algorithms are feasible to generate digital signatures according
to the German signature law -- under participation of experts from
economy and science. To review signature methods based on the
factorisation problem the GISA also co-operates with researchers of the
University of Bonn.
Further information about crypto algorithms on the web page of GISA:
   \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}
{\texttt{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}}.
}.

@@

% --------------------------------------------------------------------------
\subsubsection{One way functions}\index{One way function}
\hypertarget{OneWayFunktion2}{} 
If the key is to be impossible to determine even with knowledge of both the 
plaintext and the cipher text, we need a function that is, on the one hand, 
relatively easy to calculate -- we don't want to have problems encrypting 
messages. On the other hand, the inverse function should exist (otherwise 
information would be lost during encryption), but should be de facto 
incalculable.

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}:\\ \label{SrcArith4b}
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix D of this chapter} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

% --------------------------------------------------------------------------
\subsubsection*{Variant 1: All ASCII characters are encrypted and decrypted separately (no blocks\index{Block length} are formed).}

@@

% --------------------------------------------------------------------------
\subsubsection*{Variant 2: The ASCII characters are encrypted and decrypted two at a time as blocks.}

@@

However, it is not the 
result that is important here but rather the individual steps of the solution, that is, the explanation of the 
cryptanalysis\index{Cryptanalysis}\footnote{The method of solving the problem is outlined in the scenario of the online help to CrypTool\index{CrypTool} 
and in the presentation on the website. If anyone sends us a well prepared exact method of solving the problem, we 
would be pleased to include it in the documentation.}.

@@

\bibitem[Bourseau2002]{Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    in Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89.

@@

\bibitem[Wiles1994]{Wiles1994} \index{Wiles Andrew}
    Wiles, Andrew, \\
    {\em Modular elliptic curves and Fermat's Last Theorem}, \\
    \index{Fermat!last theorem}
    in Annals of Mathematics 141 (1995).

@@

   \item Factorisation records and challenges\index{Factorisation!factorisation records},\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.crypto-world.com/FactorWorld.html}
         {\texttt{http://www.crypto-world.com/FactorWorld.html, page by Scott Contini}} \\
         \href{http://www.loria.fr/~zimmerma/records/factor.html}
         {\texttt{http://www.loria.fr/\~{}zimmerma/records/factor.html}} \\

>>>>

\newpage
%\hypertarget{Kap_3}
\section{Introduction to Elementary Number Theory with Examples}
(Bernhard Esslinger, July 2001, Updates: Dec. 2001, June 2002, May 2003) \\

@@

Number theory or arithmetic (the emphasis here is more on the aspect of 
performing calculations with numbers) was established 
by Carl Friedrich Gauss\footnote{%
  Carl Friedrich Gauss, German mathematician and astronomer,
  Apr 30, 1777$-$Feb 23, 1855.
}
\index{Gauss Carl Friedrich} as a special mathematical discipline. Its 
elementary features include the greatest common divisor\footnote{This 
article deals with the gcd (greatest common 
divisor)\index{Gcd} in \hyperlink{Appendix_A}{Appendix A of this chapter}.}
(gcd), congruence (remainder classes), factorisation, the Euler-Fermat theorem
and primitive roots. However, the most important 
aspect is prime numbers and their multiplicative operation.

@@

\begin{center}
\fbox{\parbox{15cm}{{\em Carl Friedrich Gauss\index{Gauss Carl Friedrich}:}\\
Mathematics is the queen of sciences and number theory is the queen of mathematics.}}
\end{center}

%\begin{quote}
%{\em Carl Friedrich Gauss:}\\
%Mathematics is the queen of sciences and number theory is the queen of mathematics.
%\end{quote}

@@

Number theory arose from interest in positive whole numbers $1, 2, 3, 4, \cdots$, also referred to as the set of natural numbers 
\index{Number!natural} {\em natural numbers} $\mathbb{N}$. These are the first mathematical constructs 
used by human civilisation. According to Kronecker\footnote{Leopold
  Kronecker, German mathematician, Dec 7, 1823$-$Dec 29, 1891}\index{Kronecker Leopold}, they are a creation of God. 
In Dedekind's\footnote{Julius Wilhelm Richard Dedekind,
  German mathematician, Oct 6, 1831 $-$ Feb 12, 1916.}\index{Dedekind Julius} opinion, they are a creation of the human intellect. Dependent upon one's ideology,
this is an unsolvable contradiction or one and the same thing.

@@

If you allow numbers other than primes as factors, there are several ways of factorising integers and the {\em 
uniqueness} is lost:
$$60 = 1*60 = 2*30 = 4*15 = 5*12 = 6*10 = 2*3*10 = 2*5*6 = 3*4*5 = \cdots$$
The 1st fundamental theorem only appears to be obvious. We can construct numerous other sets of 
numbers\footnote{These sets are formed especially from the set of natural numbers. An example of this can be found in this \hyperlink{uniqueness}{script} on 
page~\pageref{thm-pz-euklid} % should be \pageref{remFundTheoOfArithm}, but
% hyperref seems to be buggy
at the end of chapter \ref{primesinmath}.
}
for which numbers in the 
set {\em cannot} be expressed 
uniquely as a product of the prime numbers of the set.

@@

{\bf Example:} \\
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that each operand is divisible by $7$: 
$69,993 = 70,000 - 7$. Therefore, the difference is also divisible by $7$.

@@

\newpage
\begin{center}
\fbox{\parbox{15cm}{
    \emph{Seneca\index{Seneca}\footnotemark:}\\
    The way of theory is long, it is short and effective by examples.
}}
\end{center}
\addtocounter{footnote}{0}
\footnotetext{%
     Lucius Annaeus Seneca, philosophical writer and poet, 4 B.\:C. $-$ 65 A.\:D.}

@@

% --------------------------------------------------------------------------
\subsubsection{Patterns and structures}
\hypertarget{Kap_3_8}{}\label{patternsandstructures}
As mathematicians investigate the structure \index{Structure} $a *x \equiv b$ mod $m$ (see \hyperlink{kap_3_5_2}{chapter 3.5.2}), so they are interested in
the structure $ x^{a} \equiv b$ mod $m.$

@@

\hypertarget{RSABeweis} Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most
common asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption. Cryptographers always associate this procedure with the
abbreviation ``RSA''. However, please note that the abbreviation may be
used in other fields to mean something completely different.}$^,$\footnote{%
In literature\index{Literature} and movies\index{Movies} not only 
classical methods are used (like the secret message solved by Sherlock Holmes
in ``The Adventure of the Dancing Men'' by Arthur Conan  
Doyle \index{Arthur Conan Doyle}), but also modern schemes. 
Examples are
% hier die items nicht einrücken ! xxxxxxxxxxxxxxx
\begin{list}{\textbullet}{\leftmargin10pt}
\item the movie ``House of Cards'', 1992, where autistic children
  communicate via primes,
\item the story ``The Dialogue of the Sisters'' by Dr.~C.~Elsner
  \index{Elsner 1999}, 1999 (included in the CrypTool package
  \index{CrypTool} as PDF file), where the sisters
  communicate confidentially using a variant of RSA.
\item the novel ``Das Museum der gestohlenen Erinnerungen (Museum of
  the stolen memories'' by Ralf Isau \index{Isau 2003}, 
  Thienemann 2003, where the last part of the oracle can only be
  solved with the joined help of the computer community.
\end{list}\vspace{-\baselineskip}\vspace{-\parskip}
}
procedure in the group $\mathbb{Z}_n^*$.

@@

{\bf 2nd comment:} \\
When it comes to signing messages, we perform the same operations but first use the secret key $d$, followed by the 
public key $e$. The RSA procedure can also be used to create digital signatures\index{Signature!digital}, because: 
$$
M \equiv (M^d)^e{\rm ~(mod~}n).
$$

@@

The security of the RSA algorithm is based on the empirical observation
that factoring large numbers is a hard problem. A module $n$ (typically,
1024 bit) can be easily constructed as the product of two large primes $p$,
$q$ (typically, 500$-$600 bit each), by calculating $n=pq$. However, it is
a hard problem to extract $p$, $q$ from $n$.  Without knowing $p$ or $q$,
the private key cannot be calculated.

@@

% --------------------------------------------------------------------------
\vskip +20pt
\paragraph{The TWIRL device} \label{TWIRLDevice} \index{TWIRL device}\mbox{}

@@

The  computations for the factorization of RSA-160 took place at the 
German Information Security Agency (GISA) in Bonn\footnote{%
Every year the GISA\index{GISA} creates a paper to describe which 
crypto algorithms are feasible to generate digital signatures according
to the German signature law -- under participation of experts from
economy and science. To review signature methods based on the
factorisation problem the GISA also co-operates with researchers of the
University of Bonn.
Further information about crypto algorithms on the web page of GISA:
   \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}
{\texttt{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}}.
}.

@@

% --------------------------------------------------------------------------
\subsubsection{One way functions}\index{One way function}
\hypertarget{OneWayFunktion2}{}%
If the key is to be impossible to determine even with knowledge of both the 
plaintext and the cipher text, we need a function that is, on the one hand, 
relatively easy to calculate -- we don't want to have problems encrypting 
messages. On the other hand, the inverse function should exist (otherwise 
information would be lost during encryption), but should be de facto 
incalculable.

@@

We will work through the example in 2 variants. The steps 1 to 3 are common for both.\par
{\bf Key generation (steps 1 to 3)}: \label{SrcArith4b} \\
{\bf 1.} $p=503, q=509 \quad (n= 256,027; \; J(n)=(p-1)(q-1)=255,016=2^3*127*251)$\footnote{%
See \hyperlink{AppArith4b}{Appendix D of this chapter} for the source code to factorise
the number $J(n)$ using Mathematica and Pari-GP.
Using CrypTool\index{CrypTool} v1.3 you can solve this with the 
{\bf Indiv.Procedures \textbackslash{} RSA Demonstration \textbackslash{} 
Factorisation of a Number}.}. \\
{\bf 2.} $e=256,027$ (should lie between $509$ and $255,016$ and must be relatively prime to $255,016$)\footnote{$e$ cannot,
therefore, be $2, 127$ or $251$ ($65537 = 2^{16}+1$).\\
In real life, $J(n)$ is not factorised but rather the Euclidean algorithm 
is used for the selected e to guarantee that ${\rm gcd}(d,J(n))=1$.}.\\
{\bf 3.} $d=231,953$ \\\strut\quad\ (since $e \equiv d^{-1}{\rm ~mod~}J(n): ~65,537*231,953 \equiv 15,201,503,761 \equiv 1
{\rm ~(mod~}67,000)$)\footnote{Other possible combinations of $(e,d)$ include: $(3, 170,011)$, $(5, 204,013)$, $(7, 36,431)$.}.

% --------------------------------------------------------------------------
\subsubsection*{Variant 1: All ASCII characters are en-/decrypted separately (no blocks\index{Block length} are formed).}

@@

% --------------------------------------------------------------------------
\subsubsection*{Variant 2: The ASCII characters are en-/decrypted two at a time as blocks.}

@@

However, it is not the 
result that is important here but rather the individual steps of the solution, that is, the explanation of the 
cryptanalysis\index{Cryptanalysis}\footnote{The method of solving the problem is outlined in the scenario of the online help to CrypTool\index{CrypTool} 
and in the presentation on the website. If anyone sends us a well prepared exact method of solving the problem, we 
would be pleased to include it in the documentation.}:

@@

\bibitem[Bourseau2002]{Bourseau2002} \index{Bourseau 2002} \index{Fox 2002}
    F. Bourseau, D. Fox, C. Thiel, \\
    {\em Vorz\"uge und Grenzen des RSA-Verfahrens},\\
    In: Datenschutz und Datensicherheit (DuD) 26/2002, pp~84-89.

@@

\bibitem[Wiles1994]{Wiles1994} \index{Wiles Andrew}
    Wiles, Andrew, \\
    {\em Modular elliptic curves and Fermat's Last Theorem}, 
    \index{Fermat!last theorem} \\
    In: Annals of Mathematics 141 (1995).

@@

   \item Factorisation records and challenges\index{Factorisation!factoring records},\\
         \href{http://www.crypto-world.com/}
         {\texttt{http://www.crypto-world.com/}} \\
         \href{http://www.crypto-world.com/FactorWorld.html}
         {\texttt{http://www.crypto-world.com/FactorWorld.html, page by Scott Contini}} \\
         \href{http://www.loria.fr/~zimmerma/records/factor.html}
         {\texttt{http://www.loria.fr/\~{}zimmerma/records/factor.html}} \\

### 2003-05-26 script/en/chapters/primes.tex
*** empty log message ***

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, besslinger@web.de, May 1999, Updates Nov. 2000, Dec. 2001, Feb. 2003)

\begin{center}
\fbox{\parbox{15cm}{
    \emph{Albert Einstein\footnotemark:}\\
    Progress requires exchange of knowledge.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  German physicist and Nobel Prize winner, March 14, 1879 $-$ April 14, 1955
}

@@

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signatures!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, s. 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fußnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich hätte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

A stronger test is provided by\index{Miller G.L.}\index{Rabin} Miller/Rabin: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

  \item \index{Catalan Eugene} \index{Catalan number} Eugene Catalan\footnote{%
Belgian mathematician, May 5, 1814 - Feb 14, 1894.

After him the so-called
{\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2  =  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.}
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

This is the \index{Prime number!theorem} \textbf{prime number theorem}. 
It was put forward by Legendre\footnote{%
Adrien-Marie Legendre, French mathematician, Sep 18, 1752$-$Jan 10, 1833.
}
\index{Legendre Adrien-Marie}
and Gauss\index{Gauss Carl Friedrich} (1777-1855) but not proved until
over 100 years later.

@@

  \item Does a polynomial\index{Polynomial} time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime?
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter\ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

      \item Does a polynomial time algorithm exist that calculates for a
            composite number $n$ a non-trivial (i.e. other than $1$ and $n$)
	    factor of $n$?\footnote{Please see chapters \ref{RSABernstein}
	    and \ref{NoteFactorisation}.}
   \end{itemize}
\end{itemize}

>>>>

\newpage
\section{Prime Numbers}
\hypertarget{Kapitel_2}{}
(Bernhard Esslinger, May 1999, Updates Nov. 2000, Dec. 2001, Feb. 2003)

\begin{center}
\fbox{\parbox{15cm}{
    \emph{Albert Einstein\footnotemark:}\\
    Progress requires exchange of knowledge.
}}
\end{center}
\addtocounter{footnote}{0}\footnotetext{%
  Albert Einstein, German physicist and Nobel Prize winner, Mar 14, 1879 $-$ Apr 14, 1955.
}

@@

Prime numbers are particularly important for 3 reasons:
\begin{itemize}
  \item In number theory, they are considered to be the basic components of
natural numbers, upon which numerous brilliant mathematical ideas are based.
  \item They are of extreme practical importance in modern
\index{Cryptography!modern} cryptography (public key \index{Cryptography!public key} cryptography). The most common public key procedure, invented at the end of
the 1970's, is \index{RSA} RSA encryption. Only using (large) prime numbers for
particular parameters can you guarantee that an algorithm is secure, both for
the RSA procedure and for even more modern procedures (digital
\index{Signature!digital} signature, elliptic curves).
  \item The search for the largest known prime numbers does not have any
practical usage known to date, but requires the best computers, is an excellent
benchmark (possibility for determining the performance\index{Performance} of computers) and leads
to new calculation methods on many computers \\ (see also:
\href{http://www.mersenne.org/prime.htm}{\tt
http://www.mersenne.org/prime.htm}).
\end{itemize}
Many people have been fascinated by prime numbers over the past two millennia.
Ambition to make new discoveries about prime numbers has often resulted in
brilliant ideas and conclusions. The following section provides an easily
comprehensible introduction to the basics of prime numbers. We will also explain
what is known about the distribution (density, number of prime numbers in
particular intervals) of prime numbers and how prime number tests work.

@@

Although there are only $92$ natural chemical elements, the number of prime
numbers is unlimited. Even the Greek, \index{Euclid} Euclid%
\footnote{Euclid,
a Greek mathematician of 4th and 3rd century B.C. He worked at the
Egyptian academy of Alexandria and wrote ``The Elements'', the most well 
known systematically textbook of the Greek mathematics.} 
knew this in the third century B.C.
\begin{theorem}[Euclid\footnote{The common usage of the term does not denote Euclid as the inventor of the theorem rather;
the true inventor is merely not as prominent. The theorem has already been distinguished
and proven in Euclid's Elements (Book IX, theorem 20). The phraseology is remarkable due to 
the fact that the word infinite is not used. The text reads as followed
$$
O\acute{\iota}~\pi\varrho\tilde{\omega}\tau o \iota~\grave{\alpha}\varrho\iota\vartheta\mu o\grave{\iota}~
\pi\lambda\varepsilon\acute{\iota}o \upsilon\varsigma~\varepsilon\grave{\iota}\sigma\grave{\iota}~
\pi\alpha\nu\tau\grave{o}\varsigma~\tau o \tilde{\upsilon}~
\pi\varrho o \tau\varepsilon\vartheta\acute{\varepsilon}\nu\tau o \varsigma~
\pi\lambda\acute{\eta}\vartheta\ o \upsilon\varsigma~
\pi\varrho\acute{\omega}\tau\omega\nu~
\grave{\alpha}\varrho\iota\vartheta\mu\tilde{\omega}\nu,
$$
the English translation of which is: the prime numbers are more than 
any previously existing amount of prime numbers.
}]\label{thm-pz-euklid} % Ende der Fußnote
% Fussnote VOR "]" in [Euclid], damit kein Leerraum vor der Fussnotennummer.
% Vorher stand da: ...(Euclid). BLANK Fussnote und das Blank stoerte.
% Nun steht die Fussnote direkt hinter "Euclid" und vor der ")".
% Eigentlich hätte ich sie gerne direkt hinter "(Euclid)", noch vor dem 
% automatisch gesetzten Punkt. 
The sequence of prime numbers does not discontinue.
Therefore, the quantity of prime numbers is infinite.
\end{theorem}
His proof that there is an infinite number of
primes is still considered to be a brilliant mathematical consideration and
conclusion today (proof by contradiction). He assumed that there is only a
finite number of primes and therefore a largest prime number. Based on this
assumption, he drew logical conclusions until he obtained an obvious
contradiction. This meant that something must be wrong. As there were no
mistakes in the chain of conclusions, it could only be the assumption that was
wrong. Therefore, there must be an infinite number of primes!

@@

A stronger test is provided by\index{Miller Gary L.}\index{Rabin Michael O.}
Miller/Rabin\footnote{
In 1976 an efficient probabilistic primality test was published by Prof. Rabin, based on a number theoretic result of Prof. Miller from the year before. \\
Prof. Miller worked at the Carnegie-Mellon University, School of Computer
Science. Prof. Rabin, born in 1931, worked at the Harvard and Hebrew University.
}%
: it is
only passed by so-called {\em strong pseudo prime numbers}. Again, there are
strong pseudo prime numbers that are not primes, but this is much less often the
case than for (simple) pseudo prime numbers. The smallest strong pseudo prime
number base $2$ is $$ 15,841 = 7 \cdot 31 \cdot 73. $$
If you test all 4 bases, $2, 3, 5$ and $7$, you will find only one strong
pseudo prime number up to $25 \cdot 10^9$, i.e. a number that passes the 
test and yet is not a prime number.

@@

  \item \index{Catalan Eugene} \index{Catalan number} Eugene Cahrles 
  Catalan\footnote{%
Eugene Charles Catalan, Belgian mathematician, May 5, 1814$-$Feb 14, 1894.\\
After him the so-called {\em Catalan numbers}
$A(n) = (1 / (n+1) ) * (2n)! / (n!)^2 $ \\
$=  1, 2, 5, 14, 42, 132, 429, 1430, 
4862, 16796, 58786, 208012, 742900, 2674440, 9694845, ... $
are named.
}
conjectured that $C_4$ is a prime:
$$
 \begin{array}{l}
C_0 = 2, \\
C_1 = 2^{C_0} - 1,  \\
C_2 = 2^{C_1} - 1,  \\
C_3 = 2^{C_2} - 1, \\
C_4 = 2^{C_3} - 1, \cdots \\
\end{array}
$$
    (see {\href{http://www.utm.edu/research/primes/mersenne.shtml}{\tt
http://www.utm.edu/research/primes/mersenne.shtml}} under Conjectures and
Unsolved Problems).

@@

This is the \index{Prime number!theorem} \textbf{prime number theorem}. 
It was put forward by Legendre\footnote{%
Adrien-Marie Legendre, French mathematician, Sep 18, 1752 $-$ Jan 10, 1833.
}
\index{Legendre Adrien-Marie}
and Gauss\index{Gauss Carl Friedrich} (1777-1855) but not proved until
over 100 years later.

@@

  \item Does a polynomial\index{Polynomial} time algorithm exist for calculating the prime factors of a number (see
    \cite[p. 167]{Klee1997})? 
    This question can be divided into the two following questions:
   \begin{itemize} 
      \item Does a polynomial time algorithm exist that decides whether a
            number is prime? \\
	    This question has been answered by the AKS algorithm\index{AKS} 
	    (see chapter \ref{PrimesinP}, ``Primes in P'': 
	    Primality testing is polynomial).

      \item Does a polynomial time algorithm exist that calculates for a
            composite number $n$ a non-trivial (i.e. other than $1$ and $n$)
	    factor of $n$?\footnote{Please compare chapters \ref{RSABernstein}
	    and \ref{NoteFactorisation}.}
   \end{itemize}
\end{itemize}
